Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-07-13 paper.tex
Describe equil f(t) plot; start describing adaptive time step

We integrate Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} with adaptive step sizes,
using the 5th-order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed},
which produces an error estimate by comparing the 4th and 5th order solutions.
\todo{update to 3rd order Runge-Kutta}
After each step is taken, the errors for each of the 9 variables of
Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} are compared.
If the maximum of these errors exceeds some absolute or relative threshold
\todo{specify ScaledAbsRel algorithm}
the step size is decreased and the step recomputed,
if it is below some threshold the next step size is increased.
% see OdeIntegrators::AdaptiveDense
% and OdeSteppers::DormandPrince5
% and OdeControllers::ProportionalIntegral_Base::StepSizeToAttempt
% and OdeErrorMeasures::ScaledAbsRel
\todo{specify}

@@

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
1) the full form  by which the solution is $f_{\rm scat}=f_{\rm AE}+f_{\rm SE}$
(labeled `scat' since we include elastic scattering), and
2) the form treating only absorption/emission interactions
in which we set $\mathscr{K}_{\rm SE}=0$ so that $f_{\rm noscat}=f_{\rm AE}$
(labeled `no scat' since we ignore elastic scattering).
There is no simple relation between the two forms for a given ray; that is
$f_{\rm scat} \neq f_{\rm noscat}+f_{\rm SE}$.
This is because $f_{\rm AE}$ is integrated using solely the absorption optical
depth in the first form,
and the absorption plus scattering optical depth in the second form.
The differences between these two methods is apparent
in Fig.~\ref{fig:cumulative_f_homogeneous}.
Before this work, the noscat form of the equations has been the standard for
ray tracing, though many authors include the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).
\todo{explain why; cite others}

@@

Fig.~\ref{fig:cumulative_f_homogeneous} displays the integrated $f$ as a
function of path length along the ray.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-cumulative_f-homogeneous}
  \caption{Cumulative distribution functions at $\varepsilon=11.25$~MeV
    in homogeneous infinite slab of Sec.~\ref{ssec:test_equilibrium}.
    The integration proceeds from left to right (backwards in time $t$).
    The two components of the scat solutions ($f_{\rm AE}$ in points,
    and $f_{\rm SE}$ in dashed lines) sum to equal the single component
    of the noscat solution ($f_{\rm AE}$ in points connected by lines).
    All solutions terminate when they achieve a total optical depth
    greater than $\tau_term=14$; in the scat cases this occurs at an
    earlier time since the total mean free paths are less than the
    absorption mean free paths (much less in the case of $\nu_x$).
    The positions of the points (only plotted for the $f_{\rm AE}$ components)
    are the time steps chosen by the adaptive time-stepping algorithm
    described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:cumulative_f_homogeneous}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure-ferr-50km}
  \caption{Relative error in integrated equilibrium distribution functions
    in the infinite homogeneous slab  described in
    Sec.~\ref{ssec:test_equilibrium}.
    Presented here are the relative differences between the final
    $f_{\rm AE}$ in the noscat case
    (see Fig.~\ref{fig:cumulative_f_homogeneous})
    and the equilibrium distribution functions given by
    Eqn.~\ref{eqn:feq}.
    As can be seen in Fig.~\ref{fig:mfps-50km} these three species
    over the given energies
    probe our numerical solution over length scales from
    100~m to $10^{5}\,{\rm km}$.
    The source of this error is the discarded boundary term
    $f_{\rm bdry}\sim10^{-6}$, described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:homogeneous_isotropic}
\end{figure}

>>>>

We integrate Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} with adaptive step sizes,
using the 3rd order Runge-Kutta algorithm
which produces an error estimate by comparing the 3rd and 2nd order solutions.
After each step is taken, the errors for each of the nine variables of
Eqns.~\ref{eqn:num_x}--\ref{eqn:num_tau} are compared.
If the maximum of these errors exceeds some absolute or relative threshold
\todo{specify ScaledAbsRel algorithm}
the step size is decreased and the step recomputed,
if it is below some threshold the next step size is increased.
% see OdeIntegrators::AdaptiveDense
% and OdeSteppers::RungeKutta3
% and OdeSteppers::DormandPrince5
% and OdeController
% and OdeControllers::ProportionalIntegral_Base::StepSizeToAttempt
% and OdeErrorMeasures::ScaledAbsRel
\todo{specify}

@@

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
1) the full form  by which the solution is $f_{\rm scat}=f_{\rm AE}+f_{\rm SE}$
(labeled \emph{scat} since we include elastic scattering), and
2) the form treating only absorption/emission interactions
in which we set $\mathscr{K}_{\rm SE}=0$ so that $f_{\rm noscat}=f_{\rm AE}$
(labeled \emph{noscat} since we ignore elastic scattering).
There is no simple relation between the two forms for a given ray; that is
$f_{\rm scat} \neq f_{\rm noscat}+f_{\rm SE}$.
This is because $f_{\rm AE}$ is integrated using solely the absorption optical
depth in the first form,
and the absorption plus scattering optical depth in the second form.
The differences between these two methods is apparent
in Fig.~\ref{fig:cumulative_f_homogeneous}.
Before this work, the \emph{noscat} form of the equations has been the standard for
\todo{explain why; cite others}
ray tracing, though many authors include the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).

@@

Fig.~\ref{fig:cumulative_f_homogeneous} displays the cumulative distribution
function integrated along the ray, for each of the three species and using
both methods, \emph{noscat} and \emph{scat}. We display
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse} in their integral form:
\begin{equation}
  \label{eqn:cumulative_f}
  f(t)
  = - \int_{0}^t dt'\,\frac{1}{p^t}e^{-\tau(0,t')}
  \mathscr{K}(t')f^{\rm eq}(t'),
\end{equation}
and the integration proceeds backwards in time, from $t=0$
\todo{could write more simply because Minkowski}
to some terminal $t<0$.
The figure shows this backwards-in-time integration proceeding left to right.

In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\todo{discuss with Matt}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-cumulative_f-homogeneous}
  \caption{Cumulative distribution functions at $\varepsilon=11.25$~MeV
    in homogeneous infinite slab of Sec.~\ref{ssec:test_equilibrium}.
    The integration proceeds from left to right (backwards in time $t$).
    The two components of the \emph{scat} solutions ($f_{\rm AE}$ in points,
    and $f_{\rm SE}$ in dashed lines) sum to equal the single component
    of the \emph{noscat} solution ($f_{\rm AE}$ in points connected by lines).
    All solutions terminate when they achieve a total optical depth
    greater than $\tau_term=14$; in the \emph{scat} cases this occurs at an
    earlier time since the total mean free paths are less than the
    absorption mean free paths (much less in the case of $\nu_x$).
    The positions of the points (only plotted for the $f_{\rm AE}$ components)
    are the time steps chosen by the adaptive time-stepping algorithm
    described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:cumulative_f_homogeneous}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure-ferr-50km}
  \caption{Relative error in integrated equilibrium distribution functions
    in the infinite homogeneous slab  described in
    Sec.~\ref{ssec:test_equilibrium}.
    Presented here are the relative differences between the final
    $f_{\rm AE}$ in the \emph{noscat} case
    (see Fig.~\ref{fig:cumulative_f_homogeneous})
    and the equilibrium distribution functions given by
    Eqn.~\ref{eqn:feq}.
    As can be seen in Fig.~\ref{fig:mfps-50km} these three species
    over the given energies
    probe our numerical solution over length scales from
    100~m to $10^{5}\,{\rm km}$.
    The source of this error is the discarded boundary term
    $f_{\rm bdry}\sim10^{-6}$, described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:homogeneous_isotropic}
\end{figure}
